---
title: "Untitled"
output:
  html_document: default
  word_document: default
  pdf_document: default
date: '2022-07-26'
---

As described in ICAO Doc 9889, aircraft emissions include emissions from the main engines, the APU
(auxiliary power units), main engine start-up and the particulate matter emissions from brake and tyre wear. In
this section, emissions from the aircraft main engines and the APU are calculated using different method
approaches and variations in performance affecting parameters.
3.1.1. Aircraft Main Engines
Emission inventories of aircraft in the vicinity of airports are traditionally calculated using ICAO engine exhaust
emission data and the ICAO reference LTO cycle, the latter sometimes adapted to airport specific taxi times.
Initially intended for certification purposes, the LTO cycle cannot sufficiently account for operational issues
(flex-rated take-off, climb profiles). Consequently, an operational aircraft LTO cycle has also been defined as
the basis for performance based modelling (Table 2).


```{r setup, include=FALSE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)

# set defaults
theme_set(theme_minimal())
```

# Inputs

## Loading data from OpenAvEm

```{r, eval = FALSE}
#openavem aircraft data
ac_classes <- read_csv("data-raw/openavem-1.0/aircraft/ac_classes.csv")
ac_icao_description <- read_csv("data-raw/openavem-1.0/aircraft/ac_icao_description.csv")
oag_to_icao <- read_csv("data-raw/openavem-1.0/aircraft/oag_to_icao.csv")
typecode_replacements <- read_csv("data-raw/openavem-1.0/aircraft/typecode_replacements.csv")

ac_classes <- ac_classes %>% rename(TYPE = typecode, MODEL_CLASS = class)
ac_icao_description <- ac_icao_description %>% rename(TYPE = Designator, DESC = Description)

# engine stuff
eng_allocations <- read_csv("data-raw/openavem-1.0/engines/eng_allocations.csv")
eng_properties <- read_csv("data-raw/openavem-1.0/engines/eng_properties.csv")

eng_allocations <- eng_allocations %>% rename(TYPE = typecode, ENG_ID = uid)
eng_properties <- eng_properties %>% rename(ENG_ID = uid)


type_engine_map <- ac_icao_description %>% 
  left_join(ac_classes, by = "TYPE") %>% 
  left_join(eng_allocations %>% select(-contains("apu")), by = "TYPE") %>% 
  left_join(eng_properties, by = "ENG_ID")

# package and save out look-up table
rq <- type_engine_map %>% 
  select(TYPE, DESC, MODEL_CLASS, ENG_ID
         , FUEL_TO = fuel_to, FUEL_CO = fuel_co
         , FUEL_APP = fuel_app, FUEL_IDLE = fuel_idle
         , FUEL_LTO = Fuel_LTO_cycle
         ) %>% 
  filter(!is.na(ENG_ID)) %>% 
  mutate( NBR_ENG = substr(DESC,2,2) %>% as.numeric() ) %>% 
  select(TYPE, DESC, CLASS, NBR_ENG, everything()) 

rq %>% write_csv("./data-analytic/type_engine_map.csv")
```


```{r}
type_engine_map <- read_csv("./data-analytic/type_engine_map.csv")

colspec <- cols_only(
     AP_C_FLTID = col_character()
    ,AP_C_REG   = col_character()
    ,ADEP_ICAO  = col_character()
    ,ADES_ICAO  = col_character()
    ,MVT_TIME_UTC = col_character()
    ,AP_C_STND = col_character()
    ,AP_C_RWY  = col_character()
    ,AP_C_ARCTYP = col_character()
    ,AC_CLASS   = col_character()
    ,BLOCK_TIME_UTC = col_character()
    ,SRC_PHASE = col_character()
    )

ds <- read_csv2(
  "~/RProjects/__DATA/EDDF/EDDF_2019_FACT.csv"
  ,trim_ws = TRUE
  , col_types = cols_only(
     AP_C_FLTID = col_character()
    ,AP_C_REG   = col_character()
    ,ADEP_ICAO  = col_character()
    ,ADES_ICAO  = col_character()
    ,MVT_TIME_UTC = col_character()
    ,AP_C_STND = col_character()
    ,AP_C_RWY  = col_character()
    ,AP_C_ARCTYP = col_character()
    ,AC_CLASS   = col_character()
    ,BLOCK_TIME_UTC = col_character()
    ,SRC_PHASE = col_character()
    )
  ) 
# strip off useless var conventions
names(ds) <- names(ds) %>% gsub(pattern = "(AP_C_)|(AC_)|(_UTC)", replacement = "",.)
ds        <- ds %>% rename(TYPE = ARCTYP)

ds <-  ds %>% mutate(MVT_TIME = lubridate::dmy_hms(MVT_TIME),BLOCK_TIME = lubridate::dmy_hms(BLOCK_TIME))
```

```{r}
acs <- ds %>% 
  group_by(TYPE) %>% 
  summarise(N = n(), ARRS = sum(SRC_PHASE == "ARR", na.rm = TRUE), DEPS = sum(SRC_PHASE == "DEP", na.rm = TRUE)) %>% 
  arrange(desc(N)) %>% 
  mutate(SHARE = N / sum(N), SHARE = prettyNum(SHARE, scientific = FALSE, digits = 3))

acs
```

```{r}
acs %>% mutate(RANK = 1:nrow(acs)) %>% 
  ggplot() +
  geom_path(aes(x = RANK, y = cumsum(SHARE)))
```

```{r}
types_not_covered <- acs %>% mutate(RANK = row_number(), CUM_SHARE = cumsum(SHARE), ENG = TYPE %in% type_engine_map$TYPE) %>% filter(ENG != TRUE)

flextable::flextable(types_not_covered)
```


```{r}
arrs <- ds %>% filter(SRC_PHASE == "ARR")
deps <- ds %>% filter(SRC_PHASE == "DEP")
```

```{r}
arr_refs <- arrs %>% 
  select(STND, RWY, ALDT = MVT_TIME, AIBT = BLOCK_TIME) %>% 
  mutate(TXIT = difftime(AIBT, ALDT, units = "min") %>% as.numeric) %>% 
  group_by(STND, RWY) %>% 
  summarise(N = n(), TXIT_REF = quantile(TXIT, probs = 0.2, na.rm = TRUE))
arr_refs

arrs_fuel <- arrs %>% 
  select(TYPE, STND, RWY, ALDT = MVT_TIME, AIBT = BLOCK_TIME) %>% 
  mutate(TXIT = difftime(AIBT, ALDT, units = "min") %>% as.numeric) %>%
  left_join(arr_refs %>% select(STND, RWY, TXIT_REF), by = c("STND","RWY")) %>% 
  left_join(type_engine_map %>% select(TYPE, NBR_ENG, ENG_ID, FUEL_IDLE), by = "TYPE") %>% 
  mutate( TXIT_FUEL     = TXIT * NBR_ENG * FUEL_IDLE * 60
         ,ADD_TXIT_FUEL = (TXIT - TXIT_REF) * NBR_ENG * FUEL_IDLE * 60)

arrs_fuel %>% select(TYPE:TXIT_REF, TXIT_FUEL, ADD_TXIT_FUEL, everything())
```

```{r}
arrs_fuel %>% mutate(DOF = lubridate::date(AIBT)) %>% select(DOF, TXIT_FUEL, ADD_TXIT_FUEL) %>% 
  pivot_longer(cols = c("TXIT_FUEL", "ADD_TXIT_FUEL"), names_to = "COMPONENT", values_to = "FUEL") %>% 
  #filter(lubridate::month(DOF) == 6) %>% 
  
  ggplot() +
    geom_col(aes(x = DOF, y = FUEL, fill = COMPONENT), position = position_stack())
```

## ACERT

```{r}
type_engine_map <- read_csv("./data-analytic/type_engine_map.csv", show_col_types = FALSE)
```


ACERT times-in-mode

```{r}
acert_tim <- tribble(
  ~"Aircraft_Group",	~"Takeoff",	~"Climb", ~	"Approach",	~ "Idle" 
  ,"Jet",	              0.7,	      2.2,	      4,           NA 
  ,"Bizjet",	          0.4,      	0.5,      	1.6,         NA
  ,"Turbo", 	          0.5,	      2.5,	      4.5,         NA
  ,"Piston",	          0.3,	       5,        	6,           NA
  ,"Helicopter",	      0,          6.5,	      6.5,      	 7.0
)
flextable::flextable(acert_tim)
``` 

EI Jet-A1 for CO2 calculations in this sheet			kg CO2/kg Jet-A1	 3,150 	 3,150 	 3,150 	V5, 2017
EI Agvas for CO2 calculations in this sheet			kg CO2/kg AVGAS	 3,253 	 3,253 	 --- 	V5, 2017

Data References:				
nnAAnnn = ICAO Engine Emission Database with ICAO UID				
A(A)nnn = Zurich Airport Engine Database with data from the Swiss FOCA and from some manufacturers				
FAEEDnnn: US FAA, AP-42 aircraft engine data base (1995)				


```{r}
library(readxl)
acert <- readxl::read_excel("./data-raw/ACERT_6.0_ACI_Public_build2216.xlsx", sheet = "EF Aircraft", skip = 2) %>% 
  mutate(ROW = row_number()) %>% 
  filter(ROW < 147)
```

# Crunching Data for BRA-EUR

load respective airport data

```{r}
apt <- "EHAM"

read_FACT <- function(.apt, .yr, .colspec, ...){
  fn <- paste0("../__DATA/", .apt, "/", .apt,"_", .yr, "_FACT.csv")
  df <- readr::read_csv2(fn, trim_ws = TRUE, col_types = .colspec)
  
  df <- df %>% mutate(across(contains("TIME_UTC"), lubridate::dmy_hms))
  names(df) <- names(df) %>% 
    gsub(pattern = "(AP_C_)|(AC_)|(_UTC)", replacement = "",.)
  df        <- df %>% rename(TYPE = ARCTYP)

  
  return(df)
}

read_multiyear_FACT <- function(.apt, .yrs, .colspec, ... ){
  dfs <- .yrs %>% 
    purrr::map_dfr(.f = ~ read_FACT(.apt, .x, .colspec))
  return(dfs)
}

# calculate REF20 and PBWG
calc_ref_txxt <- function(.df, .min_stnd_rwy = 30, .comment = "no specifics"){
    myrefs <- .df %>% 
        group_by(YEAR, ID, SRC_PHASE, STND, RWY)
    
    ref20 <- myrefs %>% 
        summarise(N = n(), VALID = sum(!is.na(TXXT)), REF_20 = quantile(TXXT, probs = 0.2, na.rm = TRUE), .groups = "drop")
    
    pbwg <- myrefs %>%
        filter(TXXT >= quantile(TXXT, probs = 0.05, na.rm = TRUE) & TXXT <= quantile(TXXT, probs = 0.15, na.rm = TRUE)) %>% 
        summarise(N_PBWG = n(), REF_PBWG = mean(TXXT), .groups = "drop")
    
    ref_comb <- left_join(ref20, pbwg, by = c("YEAR", "ID", "SRC_PHASE", "STND", "RWY")) %>% 
        mutate(OBSERVED = .comment) %>% 
        filter(VALID >= .min_stnd_rwy)
}
```

```{r}
ds2019 <- read_FACT(apt, 2019, colspec)
ds2020 <- read_FACT(apt, 2020, colspec)
ds2021 <- read_FACT(apt, 2021, colspec)

ds <- bind_rows(ds2019, ds2020, ds2021) %>% 
  distinct() %>% 
  mutate(YEAR = lubridate::year(BLOCK_TIME)
         ,TXXT = case_when(
               SRC_PHASE == "DEP" ~ difftime(MVT_TIME, BLOCK_TIME, units = "min") %>% as.numeric
              ,SRC_PHASE == "ARR" ~ difftime(BLOCK_TIME, MVT_TIME, units = "min") %>% as.numeric
              #, TRUE ~ NA
            )
         ,ID = paste(SRC_PHASE, STND, RWY, sep = "-")
         ,APT = apt
         )
refs_2019 <- ds %>%
  filter(YEAR == 2019) %>% 
  group_by(YEAR, ID, SRC_PHASE, STND, RWY) %>% 
  summarise(N = n(), VALID = sum(!is.na(TXXT)), REF_20 = quantile(TXXT, probs = 0.2, na.rm = TRUE), .groups = "drop") %>% 
  mutate(OBSERVED = "2019") %>% 
  filter(N >= 30)

refs_2020 <- ds %>% 
  filter(YEAR <= 2020) %>% 
  filter(!ID %in% refs_2019$ID) %>% 
  group_by(ID, SRC_PHASE, STND, RWY) %>% 
  summarise(N = n(), VALID = sum(!is.na(TXXT)), REF_20 = quantile(TXXT, probs = 0.2, na.rm = TRUE), .groups = "drop") %>% 
  mutate(OBSERVED = "2020, 2019") %>% 
  filter(N >= 30)

refs_2021 <- ds %>% 
  filter(YEAR <= 2021) %>% 
  filter(!ID %in% refs_2019$ID) %>% filter(!ID %in% refs_2020$ID) %>%  
  group_by(ID, SRC_PHASE, STND, RWY) %>% 
  summarise(N = n(), VALID = sum(!is.na(TXXT)), REF_20 = quantile(TXXT, probs = 0.2, na.rm = TRUE), .groups = "drop") %>% 
  mutate(OBSERVED = "2021, 2020, 2019") %>% 
  filter(N >= 20)

refs <- bind_rows(refs_2019, refs_2020, refs_2021) %>% drop_na() %>% mutate(ICAO = apt) %>% 
  select(ICAO, PHASE = SRC_PHASE, STND, RWY, N, VALID, REF_20, ID, OBSERVED)

# write_csv(refs, file = paste0("./data-analytic/", apt,"_TXXT_REF.csv"))
```

do the analysis

```{r}
apt <- "EDDF"
yrs <- c(2019,2020,2021)

ds   <- read_multiyear_FACT(apt, yrs, colspec) %>% 
  filter(CLASS != "HEL") %>% 
  mutate(DOF = lubridate::date(BLOCK_TIME), YEAR = lubridate::year(BLOCK_TIME)
         ,ID = paste(SRC_PHASE, STND, RWY, sep = "-" )
         ,TXXT = case_when(
               SRC_PHASE == "DEP" ~ difftime(MVT_TIME, BLOCK_TIME, units = "min") %>% as.numeric
              ,SRC_PHASE == "ARR" ~ difftime(BLOCK_TIME, MVT_TIME, units = "min") %>% as.numeric
            )
         )
refs <- read_csv(paste0("./data-analytic/", apt, "_TXXT_REF.csv"))

ds_fuel <- ds  %>% 
  left_join(refs %>% select(ID, REF_20), by = "ID") %>% 
  left_join(type_engine_map %>% select(TYPE, NBR_ENG, FUEL_TO, FUEL_IDLE), by = "TYPE") %>% 
  mutate(FUEL_MAP = if_else(!is.na(FUEL_IDLE), TRUE, FALSE)
         ,ADD_TXXT = TXXT - REF_20)

avg_fuel <- ds_fuel %>% 
  filter(!is.na(TXXT)) %>% 
  group_by(SRC_PHASE, CLASS) %>% 
  summarise( TXXT      = mean(TXXT, na.rm = TRUE)
            , ADD_TXXT = mean(ADD_TXXT, na.rm = TRUE)
            , NBR_ENG  = mean(NBR_ENG, na.rm = TRUE)
            , FUEL_IDLE= mean(FUEL_IDLE, na.rm = TRUE)
            , FUEL_TO  = mean(FUEL_TO, na.rm = TRUE)
            , .groups = "drop")

# join to replace NAs == fill missing == rows_patch()
ds_fuel <- ds_fuel %>% 
  rows_patch(avg_fuel, by = c("SRC_PHASE", "CLASS"))
```

ok        <- ds_fuel %>% filter(FUEL_MAP == TRUE & !is.na(ADD_TXXT))
ok_no_tx  <- ds_fuel %>% filter(FUEL_MAP == TRUE &  is.na(ADD_TXXT)) %>% 
  rows_update(avg_fuel %>% select(SRC_PHASE, CLASS, ADD_TXXT), by = c("CLASS", "SRC_PHASE"))

nok_tx   <- ds_fuel %>% filter(FUEL_MAP == FALSE) %>% 
  rows_patch(avg_fuel %>% select(SRC_PHASE, CLASS, NBR_ENG = AVG_ENG, FUEL_TO = AVG_F_TO, FUEL_IDLE = AVG_F_IDLE, ADD_TXXT), by = c("CLASS", "SRC_PHASE"))
``` 

#  mutate( ADD_TXXT = TXXT - REF_20
#         ,ADD_FUEL = ADD_TXXT * NBR_ENG * FUEL_IDLE * 60)

#write_csv(rq, file = paste0("./data-analytic/", apt, "_TXXT_FUEL.csv.gz"))
rq <- read_csv(paste0("./data-analytic/", apt, "_TXXT_FUEL.csv.gz"))
```

```{r}
taxi_fuel <- rq %>% 
  group_by(DOF, SRC_PHASE) %>% 
  summarise(  N = n()
            , VALID_TXXT = sum(!is.na(TXXT))
            , VALID_FUEL = sum(!is.na(ADD_FUEL))
            , TOT_TXXT = sum(TXXT, na.rm = TRUE)
            , TOT_REF = sum(REF_20, na.rm = TRUE)
            , TOT_ADD_TIME = sum(ADD_TXXT, na.rm = TRUE)
            ,TOT_ADD_FUEL = sum(ADD_FUEL, na.rm = TRUE)
            )
```

fuel should be corrected for missing flights

```{r}
taxi_fuel %>% 
ggplot() + 
  geom_path(aes(x = DOF, y = N - VALID_FUEL, group = SRC_PHASE)) +
  facet_wrap(.~SRC_PHASE, nrow = 2) +
  labs(title = "coverage of fleet for fuel burn estimation", y = "unknown ac types for fuel", x = NULL)
```

```{r}
p1 <- taxi_fuel %>% group_by(DOF) %>% 
  summarise(MVTS = sum(N, na.rm = TRUE)) %>% 
  ggplot() + 
    geom_line(aes(x = DOF, y = MVTS)) +
    labs(x = NULL, y = "movements")

p2 <- taxi_fuel %>% 
  ggplot() +
  geom_path(aes(x = DOF, y = TOT_ADD_FUEL, colour = SRC_PHASE)) +
  scale_y_continuous(labels = scales::label_number(suffix = " k", scale = 1e-3)) +
  labs(x = NULL, y = "total additional fuel burn (kt)"
       ,colour = NULL) +
  theme(legend.position = c(0.8, 0.8))

p3 <- taxi_fuel %>% 
  ggplot() +
  geom_col(aes(x = DOF, y = TOT_ADD_FUEL, colour = SRC_PHASE)) +
  scale_y_continuous(labels = scales::label_number(suffix = " k", scale = 1e-3)) +
  labs(x = NULL, y = "total additional fuel burn (kt)"
       ,colour = NULL) +
  theme(legend.position = c(0.8, 0.8))

p1 / p2 + 
  plot_layout(heights = c(1,4)) +
  plot_annotation(title = "EDDF - daily movements and additional fuel burn")
```


```{r}
tmp <- taxi_fuel %>% 
  group_by(DOF) %>% 
  summarise(TOT_ADD_FUEL = sum(TOT_ADD_FUEL)) %>% 
  mutate(CUM_TOT_ADD_FUEL = cumsum(TOT_ADD_FUEL))

tmp_max_year <- tmp %>%
  group_by(YEAR = lubridate::year(DOF)) %>% 
  summarise(MAX = max(CUM_TOT_ADD_FUEL)) %>% 
  mutate(NEXT_YEAR = YEAR + 1) %>% 
  bind_rows(tibble(YEAR = 2018, MAX = 0, NEXT_YEAR = 2019))

tmp <- tmp %>% mutate(YEAR = lubridate::year(DOF)) %>% 
  left_join(tmp_max_year %>% select(-YEAR) %>% rename(YEAR = NEXT_YEAR)
            , by = "YEAR") %>% 
  mutate( DOY = lubridate::yday(DOF)
         ,ANNUAL_ADD_FUEL = CUM_TOT_ADD_FUEL - MAX)
  
tmp %>% 
  ggplot() + 
    geom_line(aes(x = DOF, y = CUM_TOT_ADD_FUEL)) +
    scale_y_continuous(labels = scales::label_number(suffix = " M", scale = 1e-6)) +
    labs(x = NULL, y = "additional taxi fuel burn (Mt)", title = "cummulative additional fuel burn since 1. Jan 2019")
```

```{r}
tmp %>% mutate(FAKE_DOF = lubridate::`year<-`(DOF,2000)) %>% 

  ggplot() + 
    geom_line(aes(x = FAKE_DOF, y = ANNUAL_ADD_FUEL, group = YEAR, colour = as.factor(YEAR))) +
    scale_x_date(breaks = "1 month", date_labels = "%b") +
    scale_y_continuous(labels = scales::label_number(suffix = " M", scale = 1e-6)) +
    labs(x = NULL, y = "cumulative annual additional fuel"
         , colour = NULL) +
    theme(legend.position = c(0.1, 0.8))
```

```{r}

```

