---
title: "Fuel Burn Estimation for Operational Performance Benchmarking"
date: "July 2022"
author:
  - first_name: "Rainer"
    last_name: "Koelle" 
    #url: https://example.com/norajones
    affiliation: "EUROCONTROL, Performance Review Unit"
    affiliation_url: https://example.com/spacelysprokets
    orcid_id: 0000-0003-2810-1404
citation_url: https://rstudio.github.io/distill
bibliography: ANS_Perf_Book.bib
---

First stab at an article summarising this research.    
This is **work in progress**. Please consult this page for updates in the future.

```{r, include=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

# --------- set defaults -----------
knitr::opts_chunk$set(
  echo = FALSE)
ggplot2::theme_set(theme_minimal())
```

# Abstract{-}

# Introduction

Across the globe states have implemented *Green Deal* policies and priorities. 
Air transportation accounts for about 2-3% of annual CO2 emissions, however, the total amount of CO2 emissions constantly increased pre-COVID.

(ref:iataco2) Annual emissions by commercial aircraft 2004-2021

```{r iataco2, fig.cap="(ref:iataco2)"}
#https://www.iata.org/en/iata-repository/publications/economic-reports/airline-industry-economic-performance---2015-end-year---tables/
#https://www.iata.org/en/iata-repository/publications/economic-reports/airline-industry-economic-performance---december-2019---data-tables/
#https://www.iata.org/en/iata-repository/publications/economic-reports/airline-industry-economic-performance---june-2022---data-tables/
iata_co2 <- 
  tribble(
    ~YEAR, ~CO2 
    ,2004, 628
    ,2005, 651
    ,2006, 658
    ,2007, 675
    ,2008, 669 
    ,2009, 634
    ,2010, 665
    ,2011, 685
    ,2012, 693
    #,2014, 709    # discrepancy across stats sheet
    ,2014, 733
    ,2015, 774
    ,2016, 812
    ,2017, 860
    ,2018, 905
    ,2019, 936
    ,2020, 495
    ,2021, 577
  )

iata_co2 %>% 
  ggplot() +
  geom_line(aes(x = YEAR, y = CO2, group = "iata")) +
  labs(x = NULL, y = "Commercial airline CO2 (Mt)")
``` 

Fig \@ref(fig:iataco2) shows the total emissions of commercial flights as reported by IATA [@iataIndustryStatisticsFact2019].

Levers for fuel reduction:

* operational efficiency
* market-based measures
* sustainable aviation fuel
* new aircraft propulsion and airframes

A wider use and pick-up of sustainable aviation fuel, and new aircraft propulsion technologies or aircraft design requires further research.
Despite the introduction of an initial market-based mechanism, immediate action to curb fuel burn and CO2 emissions rests with improvements of operational efficiency.

ICAO promotes a performance-based approach.
The Global Air Navigation Plan (GANP) proposes indicators for regional benchmarking [@icao_doc9750_2019].
However, there is no detailed guidance on how to measure *additional fuel burn*.

The contribution of this paper are as follows:

* development of supporting material to estimate fuel burn / CO2 emissions as part of the operational performance benchmarking
* development and application of a the approach to additional time in terminal airspace and surface movement phases (i.e. taxi-in and taxi-out)

# Background

## ICAO GANP Key Performance Indicators

>provide overview of GANP, KPIs
>zoom in on ASMA, taxi-in/taxi-out

# Conceptual Approach

Operational inefficiencies typically increase the aircraft flying time (i.e. airborne and surface movement times) and , thus, engine running time.
Engine time is directly linked to fuel burn and associated emissions and pollutants.
In that respect, inefficiencies contribute to the ditremental effect of excessive emissions to climate change.

Aircraft engines produce various types of emissions. These emissions include carbon dioxide $CO_2$, water vapour ${H_2}O$, nitrogen oxides
nitrogen oxides (NOX), carbon monoxide (CO), unburned hydrocarbons (HC), sulfur oxides (SOX), particulate matter (PM), and other trace compounds. 
About 70 percent of aircraft emissions are CO2; followed by H2O at slightly less than 30% while the rest of the pollutant represent less than 1 percent each [source]. 
CO2 and H2O form significant amount of GHG emissions that can trigger climate change while NOX, CO, HC, SOX and particulate matter are always associated with air quality and subsequently public health.

This research focuses on fuel burn and CO2 emissions stemming from operational inefficiencies during the arrival and surface movement phase of flight.


# Method and Materials

## Data

* analytic data for ASMA, taxi-in/out
* fuel burn estimate data


# Results and Discussion

# Conclusions

